---
title: "Unpacking why almost half of COVID-19 cases in Toronto have no information about their source"
subtitle: "Investigating possible data collection biases using patient demographics"
author: 
  - Sidharth Gupta
thanks: "Code and data are available at: https://github.com/sidguptacode/COVID_Sources_Analysis"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Data about the sources of COVID-19 infection can be used to guide epidemic and lockdown policies as the pandemic evolves; however almost half of COVID-19 cases in Toronto have no information about their source. In this paper we analyze the patient demographics of these cases with no information, and when they occur. We identify both groups of patients and time-points in the pandemic where there is a large proportion of cases with no source information. Our results can be used to improve the current methods for collecting data about COVID-19 cases in Toronto and other regions so that more sources of infection can be identified."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("bookdown")
library(ggmap)
library(dplyr)
library(forcats)
library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(wrapr)
```

# Introduction

The COVID-19 pandemic has made a major global and societal impact. From its beginning to this time of writing, there have over 390 million positive recorded cases of COVID-19. The virus has a contagion that is exponential in nature, and spreads easily in most indoor and close-contact settings. To counteract the rapid and dangerous spread of COVID-19, lockdown and social isolation policies are regularly reviewed and enforced. These lockdowns are effective in controlling the spread of COVID-19, however they have shown to create many societal, economic, and global mental health consequences. For instance, [@leandnguyen] found that lockdowns cause people to experience adverse psychological events, such as anxiety, disinterest, depression, and an overall reduction in mental health. [@brunolarue] analyzed that lockdowns in Canada impact low-income workers and vulnerable communities significantly with employment instability, and  [@kaurvibhu] discover that lockdowns reduce the ability to provide services for social causes such as homelessness. Many new methods have been introduced for improving lockdown design, such as modelling pandemic mobility with data [@mobility1],  and developing public health messaging systems [@Block2020]. These methods stem from the prevalent topic in the literature that lockdown strategies should be smart -- that is, they should minimize isolation consequences and maximize virus spread containment [@Wang2021, @Olivier2020]. Datasets that identify outstanding regions for COVID-19 spread are vital, because they can identify areas that most benefit from lockdown, and areas that least benefit. As of the time of this writing, the City of Toronto has been recording this kind of data from positive COVID-19 cases since the start of the pandemic. Each recorded COVID-19 case has a field that identifies the source of infection; however 42% of cases in this dataset have no information entered for this field. Such a large proportion of missing data about the source of infection significantly reduces this dataset's utility for aiding smart lockdown designs. Additionally, this large proportion of missing data suggests that there are flaws within the data collection process itself. Because this COVID-19 data is collected weekly and is used for high-stakes decision making, finding flaws in data collection and addressing them quickly is pivotal.

In this paper, we thoroughly describe how this dataset is collected, and propose three biases that can explain why data for the source of infection is missing. We affirm the plausibility of these proposed by visualizing the data in three perspectives. The first perspective visualizes how the proportion of cases with no source information is related to the different months in the pandemic's lifetime. The second perspective illustrates the age groups of individuals who have a high proportion of cases with this field missing. Finally, we cross reference the neighbourhoods in this dataset with another dataset to find their income index values, and observe that neighbourhoods with a high proportion of missing fields in their data have a lower income index than neighbourhoods with a lower proportion. With data to support the plausiblity of these biases, we hope that our results can be used to further investigate the data collection process of COVID cases in Toronto and reduce the proportion of records with this field missing.

# Data

The dataset of COVID-19 cases in Toronto is provided by the Open Data Toronto portal [@opendatatoronto]. This dataset is curated by the City of Toronto [@opendata_covid] and it is refreshed with new records on a weekly bases. The raw data comes from The Public Health Case and Contact Management (CCM) group, who collect COVID-19 case and outbreak from all of Ontario's Public Health Units at 1:00 P.M Eastern time each day [@covid]. The first row of the curated dataset are shown in Table \@ref(tab:first-row) and Table \@ref(tab:first-row-continued); some fields are not shown in these tables due to their rare occurrence. The focus of our paper will be the header for "Source of Infection". As stated in the dataset's technical guide [@covid_technical_guide], this field is determined by a public health investigator's assessment. If this assessment is absent, then other data fields may have been used to estimate the source of infection, such as if there a recorded household positive case. A breakdown of responses for this field is shown in Table \@ref(tab:source-dist-table). 

```{r, echo = FALSE, message=FALSE}

# Run the RScript which imports required helper functions for this project.
source("../../scripts/00-helpers.R", local = knitr::knit_global())

# Run the RScript for initial data cleaning. This script gives us the following dataframes to work with:
# covid_cases_clean: all recorded COVID-19 cases
# cases_with_info: all records in covid_cases_clean with information about their source of infection.
# cases_no_info: all records in covid_cases_clean without information about their source of infection.
# income_data: a dataframe that maps neighbourhoods to low-income index.
source(here::here("scripts/02-data_cleaning.R"), local = knitr::knit_global())
```

```{r first-row, echo = FALSE}
knitr::kable(head(covid_cases[3:7], n=1),
  booktabs=TRUE,
  caption="One record in the COVID-19 cases data",
  digits=2
)
```

```{r first-row-continued, echo = FALSE}
knitr::kable(head(covid_cases[8:12], n=1),
  booktabs=TRUE,
  caption="One record in the COVID-19 cases data, continued",
  digits=2
)
```

```{r source-dist-table, echo = FALSE, message=FALSE}
kable(covid_cases_all_sources,
  booktabs=TRUE,
  caption="Distribution of each response to the Source of Infection field.",
  col.names = c("Source of Infection", "Proportion of response against all cases"),
  digits=2
)
```

The described data collection process has many venues for biases to be introduced, and we will discuss three such biases. The first centers around traffic in different Public Health Units (PHUs). The recorded COVID-19 cases come from individual PHUs across the neighbourhoods of Toronto, and in an unbiased dataset, each COVID-19 case would contain equal information even if it was processed in a different PHU. However, different PHUs receive different amounts of traffic, especially during months when the pandemic shows a high surge of cases. Labour shortages in healthcare is a well-known issue in metropolitan cities that has been severely affected by COVID-19 [@Mascha2020], and as such different PHUs may not have equal capacities to handle noise in data collection, or even collect fields like the source of infection at all. A second venue for bias comes with the process of identifying the sources of COVID-19 infection through human interactions. Symptoms for COVID-19 can occur several days after the virus is initially contracted, and so asking a patient where they think the source of infection depends entirely on their own memory and comfort levels. Several studies in cognitive psychology [@Bard20152016] have shown that human memory is unreliable and irrational, and distrust can sway decision making [@distrust]. The third bias digs deeper in the issues of trust; even if a patient is confident about the source of their infection, they may not feel comfortable sharing that information. Some lockdown policies have legal consequences if they are not followed, and so some patients may fear the potential consequences that come with sharing their information. We would expect to see higher levels of distrust in low-income neighbourhoods, because those are often populated by people who are most vulnerable to legal consequences.

We will now describe our data analysis pipeline for investigating these proposed biases, implemented using the R programming language [@citeR]. We start by exploring how surges in cases influence the proportion of cases with no information about the source. We load the whole dataset using the R Tidyverse library [@thereferencecanbewhatever], and clean the table headers using the R janitor library [@janitor], the forcats library [@forcats], and the wrapr libarry [@wrapr]. Next, we use the R tidyr library [@tidyr] and dplyr library [@dplyr] to select the records in the dataset that contain no information about the source, and store those records in a new smaller data frame. In both data frames, we use the lubridate library [@lubridate] to count the number of records that occur in a given month. We then divide the counts of the smaller data frame by the total data frame to get a proportion of cases per month where there is no information about the source. We normalize the proportion values in the smaller data frame, and the count values in the total dataframe, and create a stacked bar plot of those normalized values using the ggmap library [@ggmap]. Our resulting stacked bar chart is shown in Figure \@ref(fig:cases-timepoints-plot), compiled using the knitr library [@knitr]. The results in Figure \@ref(fig:cases-timepoints-plot) show a surge in proportions right before a surge in case counts, indicating that right before a peak in cases, participants are unlikely to give information about their source of infection. This can make sense, as the time before these peaks brings upon a lot of uncertainty for many people, and also for many healthcare workers who must prepare for a new forcasted demand of cases.

```{r, echo = FALSE, message=FALSE}
source(here::here("scripts/03-monthly_cases_analysis.R"), local = knitr::knit_global())
```

```{r cases-timepoints-plot, echo = FALSE, message=FALSE, fig.cap="Visualizing normalized total COVID-19 case counts (in blue) with normalized proportions of cases with no information (in red).", warning=FALSE}
normalize_dfs_and_barplot(counts_for_reported_date, "n", reported_data_noinfo_prop_sorted, "prop", "reported_date", "Month and year", " Normalized case count / proportion", "Visualizing proportions against time", TRUE, TRUE) 
```

Our second analysis visualizes the proportion of source information for different age groups. Using the with records containing no source information, we count the number of records belonging to each age group, and again divide the results with the total case counts for each age group, visualizing the results in Figure \@ref(fig:age-group-graph). If we exclude the group of ages 19 years and younger, we can see that the proportion of cases with no source monotonically decreases as the age groups increase. This can provide support to the argument of distrust as a data bias, for older age groups stand more to gain from telling information about the source, because that information can be used for their treatment of the virus. Younger groups are less affected by the virus physically, and so they may not require the treatment benefits from telling the truth.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
source(here::here("scripts/04-age_analysis.R"), local = knitr::knit_global())
```

```{r, age-group-graph, echo = FALSE, message=FALSE, warning=FALSE, fig.cap="Visualizing the proportion of cases with no source per age group."}
single_df_barplot(age_proportions_against_total, "age_group", "prop", "Age Group", "Proportion of cases with no source", "Age demographics of cases with no source")
```

Our final analysis looks at the neighbourhoods in our proportion data frame. We take the top 30 neighbourhoods with highest proportions of no source information, and cross-reference the income indices of thos neighbourhoods. The results are visualized in Figure \@ref(fig:top-no-info-plot). The same process is done for the bottom 30 neighbourhoods -- neighbourhoods that often provide source information -- against their income indices, shown in Figure \@ref(fig:bot-no-info-plot). In Figure\@ref(fig:top-no-info-plot), we can see that many of the neighbourhoods have an income index below 0.25. We do not observe this behaviour in Figure \@ref(bot-no-info-plot), for most of the normalized income indices are close to 0.5. This finding indeed suggests that the economic status of neighbourhoods is indeed a factor for collecting data about the source of infections, and should be taken into consideration at PHUs in those regions.

```{r, echo = FALSE, message=FALSE}
source(here::here("scripts/05-neighbourhood_income_analysis.R"), local = knitr::knit_global())
```

```{r top-no-info-plot, echo = FALSE, message=FALSE, fig.cap="Comparing the normalized income indices (in red) of the top 30 neighbourhoods with high proportions (in blue) of no source information.", warning=FALSE}
normalize_dfs_and_barplot(top_noinfo_neighbs, "prop", income_data_of_topneighbs_df, "income_index", "neighbourhood_name","Neighbourhood", "Normalized income index / proportion", "Top-30 neighbourhoods responding with no source against income", FALSE, FALSE) 
```

```{r bot-no-info-plot, echo = FALSE, message=FALSE, warning=FALSE, fig.cap="Comparing the income indices (in red) of the loweest 30 neighbourhoods with proportions (in blue) of no source information."}
normalize_dfs_and_barplot(bottom_noinfo_neighbs, "prop", income_data_of_botneighbs_df, "income_index", "neighbourhood_name", "Income Index", " Neighbourhood", "Bottom-30 neighbourhoods responding with no source against income", FALSE, FALSE) 
```


# References


